### 2023-02-28 calculus.tex
Use American spelling for synthesize and analyze

Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
judgments. \emph{Type synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under
the typing context $\ctx$, the expression $e$ synthesises or infers the type $\TMV$. \emph{Type
analysis}, written $\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
where an expression of type $\TMV$ is expected. We may give these typing semantics traditionally,
without considering error localization and recovery.

@@

The marked language possesses its own typing semantics, also formulated bidirectionally. We write
$\ctxSynType{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ for analysis.
Finally, marking is bidirectional as well. The synthetic marking judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that $\EMV$ is marked into $\ECMV$, which analyses against $\TMV$.

@@

Furthermore, marking should preserve the syntactic structure of programs, modulo error marks. Thus,
we define a notion of \emph{mark erasure}, whose definition is given
\cref{fig:calculus-mark-erasure}, which converts marked expressions back into unmarked expression by
removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
\emph{well-formedness} criterion marking.
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, if an unmarked expression is marked into some marked expression, erasing error marks
from the marked expression gives the original unmarked expression. We also ensure that marking gives
well-typed marked expressions; indeed, their types should match the types synthesised or analysed
against in marking. % TODO more intuition for this theorem

@@

In addition expressions, the type of both operands should be $\TNum$. In a bidirectional system, we
thus analyse the operands against $\TNum$, giving the following rule:
%
\begin{mathpar}
  \judgment{
    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\
    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
  }{
    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
  }{USPlus}
\end{mathpar}

@@

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MALam2}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analysed against the unknown type in \textsc{IALam2}.

@@

Finally, we consider the subsumption rule in typing unmarked expressions, which states that if an
expression synthesizes a type, it may also be analysed against that type or any consistent type.
\[%
  \judgment{
    \ctxSynType{\ctx}{\EMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaType{\ctx}{\EMV}{\TMV}
  }{UASubsume}
\]%
We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
conditionals, the only ones with explicit analysis rules. In other words, we restrict
subsumption to be the analytic rule of ``last resort''. This is critical in establishing that
marking and typing are deterministic (discussed later in \cref{thm:calculus-marking-unicity}).

@@

However, it may be the case that the expression is inconsistent with the expected type. Recalling
the example of \cref{fig:calculus-examples-inconsistent-types}, subsumption would used to analyse
$\ETrue$ against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$,
and $\ETrue$ should be marked with an inconsistent type mark. Hence, the following final set of
rules is required to handle such cases:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\\\
    \subsumable{\EMV}
  }{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
  }{IAInconsistentTypes}

>>>>

Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
judgments. \emph{Type synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under
the typing context $\ctx$, the expression $e$ synthesizes or infers the type $\TMV$. \emph{Type
analysis}, written $\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
where an expression of type $\TMV$ is expected. We may give these typing semantics traditionally,
without considering error localization and recovery.

@@

The marked language possesses its own typing semantics, also formulated bidirectionally. We write
$\ctxSynType{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ for analysis.
Finally, marking is bidirectional as well. The synthetic marking judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that $\EMV$ is marked into $\ECMV$, which analyzes against $\TMV$.

@@

Furthermore, marking should preserve the syntactic structure of programs, modulo error marks. Thus,
we define a notion of \emph{mark erasure}, whose definition is given
\cref{fig:calculus-mark-erasure}, which converts marked expressions back into unmarked expression by
removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
\emph{well-formedness} criterion marking.
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, if an unmarked expression is marked into some marked expression, erasing error marks
from the marked expression gives the original unmarked expression. We also ensure that marking gives
well-typed marked expressions; indeed, their types should match the types synthesized or analyzed
against in marking. % TODO more intuition for this theorem

@@

In addition expressions, the type of both operands should be $\TNum$. In a bidirectional system, we
thus analyze the operands against $\TNum$, giving the following rule:
%
\begin{mathpar}
  \judgment{
    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\
    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
  }{
    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
  }{USPlus}
\end{mathpar}

@@

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MALam2}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analyzed against the unknown type in \textsc{IALam2}.

@@

Finally, we consider the subsumption rule in typing unmarked expressions, which states that if an
expression synthesizes a type, it may also be analyzed against that type or any consistent type.
\[%
  \judgment{
    \ctxSynType{\ctx}{\EMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaType{\ctx}{\EMV}{\TMV}
  }{UASubsume}
\]%
We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
conditionals, the only ones with explicit analysis rules. In other words, we restrict
subsumption to be the analytic rule of ``last resort''. This is critical in establishing that
marking and typing are deterministic (discussed later in \cref{thm:calculus-marking-unicity}).

@@

However, it may be the case that the expression is inconsistent with the expected type. Recalling
the example of \cref{fig:calculus-examples-inconsistent-types}, subsumption would used to analyze
$\ETrue$ against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$,
and $\ETrue$ should be marked with an inconsistent type mark. Hence, the following final set of
rules is required to handle such cases:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\\\
    \subsumable{\EMV}
  }{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
  }{IAInconsistentTypes}
